---
title: "Imputation of height"
author: "David M Hannon"
output: 
  html_document:
    toc: true
    toc_depth: 6
    toc_float: true
    theme: journal
    number_sections: false

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readxl)
library(DataEditR)
library(tableone)
library(gt)
library(gtsummary)
library(mice)
set.seed(1)
rm(list = ls())
```

## Introduction

The .xlsx master file is read in.

```{r read_in_xlsx, echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
path <- ('/Users/davidhannon/Documents/02. Medicine/Med_Programming/00. Patient DB/around_proning_for_ML/around_proning.xlsx')

population_prone <- read_xlsx(path = path, guess_max = 100) %>% 
  select(
    patient_id,
    gender,
    age,
    weight,
    height_recorded,
    height,
    total_proning,
    apache_ii,
    location_preadmission,
    los,
    outcome
  ) %>% 
  distinct()
```

We now have a dataframe of the entire population who are in the study (n = `r nrow(population_prone)`). Of these, the patient height could be located for `r sum(population_prone$height_recorded == 'T')`, and not recorded or locatable for `r sum(population_prone$height_recorded == 'F')`.

We must initially replace all heights that are ICNARC standard values with NA. We can then see what missing values we are dealing with in a visualisation.

```{r replace_na, echo=TRUE, results='hide'}
population_prone$height[population_prone$height_recorded == "F"] <- NA
md.pattern(population_prone, rotate.names = T)

```

\

We will ensure the variables are encoded correctly.

```{r include=FALSE}

population_prone$gender <- as.factor(population_prone$gender)
population_prone$age <- as.integer(population_prone$age)
population_prone$total_proning <- as.integer(population_prone$total_proning)
population_prone$apache_ii <- as.integer(population_prone$apache_ii)
population_prone$location_preadmission <- as.factor(population_prone$location_preadmission)
population_prone$outcome <- as.factor(population_prone$outcome)

```

And then run the MICE imputation

```{r include=FALSE}

prone_impute_raw <- select(population_prone, -height_recorded)
prone_model <- mice(data = prone_impute_raw, m = 100, method = 'pmm')
rm(prone_impute_raw)
```

We now have imputed values for the `r sum(population_prone$height_recorded == 'F')` heights that were  missing. Let's create a new dataframe of imputed values. We will also add ina variable to show if it is an imputed value or not.

```{r}
# turn the imputed data into a tibble
prone_demo_imputed <- as_tibble(complete(prone_model))

# get raw demographics data frame ready for the join
population_prone <- select(population_prone, patient_id, height_recorded)

# join
prone_demo_imputed <- 
  left_join(x = prone_demo_imputed,
           y = population_prone,
           by = "patient_id")

# add var to indicate if height was imputed
prone_demo_imputed$height_imputed <- prone_demo_imputed$height_recorded != 'T'
select(prone_demo_imputed, -height_recorded)

```

Now create a summary of the new demographics

```{r}
# theme_gtsummary_journal(journal = "jama")
# Setting theme `JAMA`

theme_gtsummary_compact()
# Setting theme `Compact`

table_1 <- prone_demo_imputed %>%
  select(-c(height_recorded, height_imputed, patient_id)) %>% 
  tbl_summary(
    statistic = list(all_continuous() ~ "{mean} ({sd})",
                     all_categorical() ~ "{n} / {N} ({p}%)"),
    digits = all_continuous() ~ 1,
    label = list(gender ~ "Gender", 
                 age ~ "Age (years)",
                 weight ~ "Weight (kg)",
                 height ~ "Height (cm)",
                 #bmi ~ "BMI",
                 apache_ii ~ "Apache II",
                 total_proning ~ "Total proning sessions",
                 location_preadmission ~ "Admitting location",
                 los ~ "Length of stay (days)",
                 outcome ~ "Outcome")
  ) %>% 
  # modify_caption("**Table 1. Patient Characteristics**") %>% 
  bold_labels() %>% 
  modify_footnote(
    all_stat_cols() ~ "Mean (SD) or Frequency (%)
    "
  ) %>% 
  as_gt() %>% 
  gt::tab_source_note(gt::md("some heights imputed"))

table_1

```

